Staged Abstract Interpreters

### 2019-08-16 paper/20_prelim.tex
update and spaces
typo

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
  trait Monad[M[_]] {                                  
    def pure[A](a: A): M[A]                            
    def flatMap[A,B](ma: M[A])(f: A => M[B]): M[B]     
  }                                                    
    \end{lstlisting}
    \caption{trait \texttt{Monad}} \label{fig:monad}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \begin{lstlisting}
trait MonadOps[M[_], A] {
  def map[B](f: A => B): M[B]
  def flatMap[B](f: A => M[B]): M[B]
}
    \end{lstlisting}
    \caption{trait \texttt{MonadOps}} \label{fig:monadops}
  \end{subfigure}
\end{figure}

@@

Other primitive operations provide basic functionality for the
interpreter.  The method \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{num} and \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{close} deal with primitive values,
which lift literal terms (e.g., lambdas) to our value representation
(e.g., closures).  The method \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{get} retrieves the value of an
identifier \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{x}, through the environment and store. Conditionals and
arithmetic is handled by \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{br0} and \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{arith}, respectively. The method
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{ap_clo} is used for function applications: it takes a function value
and an argument value. Note that the \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Env}, \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Store}, and \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Value} are
all annotated by \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{R[_]} since they can potentially be next-stage
values when acting as compilers.
\begin{lstlisting}
  def num(i: Int): Ans
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(rator: R[Value], rand: R[Value]): Ans
\end{lstlisting}

\begin{figure}[h!]
  \centering
  \begin{lstlisting}
          def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
            case Lit(i) => num(i)                   case Let(x, rhs, e) => for {
            case Var(x) => for {                      v  <- ev(rhs)
              ρ <- ask_env                            ρ  <- ask_env
              σ <- get_store                          α  <- alloc(x)
            } yield get(σ, ρ, x)                      _  <- set_store(α → v)
            case Lam(x, e) => for {                   rt <- local_env(ev(e))(ρ + (x → α))
              ρ <- ask_env                          } yield rt
            } yield close(ev)(Lam(x, e), ρ)         case Aop(op, e1, e2) => for {
            case App(e1, e2) => for {                 v1 <- ev(e1)                                               
              v1 <- ev(e1)                            v2 <- ev(e2)
              v2 <- ev(e2)                          } yield arith(op, v1, v2)
              rt <- ap_clo(ev)(v1, v2)              case Rec(x, rhs, e) => for {
            } yield rt                                α  <- alloc(x)
            case If0(e1, e2, e3) => for {             ρ  <- ask_env
              cnd <- ev(e1)                           v  <- local_env(ev(rhs))(ρ + (x → α))
              rt  <- br0(cnd, ev(e2), ev(e3))         _  <- set_store(α → v)
            } yield rt                                rt <- local_env(ev(e))(ρ + (x → α))
                                                    } yield rt                    
          }
  \end{lstlisting}
\caption{The generic interpreter,
  shared by the unstaged/staged + concrete/abstract interpreter.}
\label{fig:shared_int}
\end{figure}

>>>>

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
  trait Monad[M[_]] {                                  
    def pure[A](a: A): M[A]                            
    def flatMap[A,B](ma: M[A])(f: A => M[B]): M[B]     
  }                                                    
    \end{lstlisting}
    \caption{trait \texttt{Monad}} \label{fig:monad}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.4\textwidth}
    \begin{lstlisting}
trait MonadOps[M[_], A] {
  def map[B](f: A => B): M[B]
  def flatMap[B](f: A => M[B]): M[B]
}
    \end{lstlisting}
    \caption{trait \texttt{MonadOps}} \label{fig:monadops}
  \end{subfigure}
\end{figure}
\vspace{-1em}

@@

Other primitive operations provide basic functionality for the
interpreter.  The method \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{num} and \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{close} deal with primitive values,
which lift literal terms (e.g., lambdas) to our value representation
(e.g., closures).  The method \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{get} retrieves the value of an
identifier \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{x}, through the environment and store. Conditionals and
arithmetic are handled by \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{br0} and \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{arith}, respectively. The method
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{ap_clo} is used for function applications: it takes a function value
and an argument value. Note that the \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Env}, \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Store}, and \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Value} are
all annotated by \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{R[_]} since they can potentially be next-stage
values when acting as compilers.
\begin{lstlisting}
  def num(i: Int): Ans
  def get(σ: R[Store], ρ: R[Env], x: Ident): R[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(rator: R[Value], rand: R[Value]): Ans
\end{lstlisting}

\begin{figure}[h!]
  \centering
  \begin{lstlisting}
          def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
            case Lit(i) => num(i)                   case Let(x, rhs, e) => for {
            case Var(x) => for {                      v  <- ev(rhs)
              ρ <- ask_env                            ρ  <- ask_env
              σ <- get_store                          α  <- alloc(x)
            } yield get(σ, ρ, x)                      _  <- set_store(α → v)
            case Lam(x, e) => for {                   rt <- local_env(ev(e))(ρ + (x → α))
              ρ <- ask_env                          } yield rt
            } yield close(ev)(Lam(x, e), ρ)         case Aop(op, e1, e2) => for {
            case App(e1, e2) => for {                 v1 <- ev(e1)                                               
              v1 <- ev(e1)                            v2 <- ev(e2)
              v2 <- ev(e2)                          } yield arith(op, v1, v2)
              rt <- ap_clo(ev)(v1, v2)              case Rec(x, rhs, e) => for {
            } yield rt                                α  <- alloc(x)
            case If0(e1, e2, e3) => for {             ρ  <- ask_env
              cnd <- ev(e1)                           v  <- local_env(ev(rhs))(ρ + (x → α))
              rt  <- br0(cnd, ev(e2), ev(e3))         _  <- set_store(α → v)
            } yield rt                                rt <- local_env(ev(e))(ρ + (x → α))
                                                    } yield rt                    
          }
  \end{lstlisting}
\caption{The generic interpreter,
  shared by the unstaged/staged + concrete/abstract interpreter.}
\label{fig:shared_int}
\end{figure}
\vspace{-1em}

### 2019-08-16 paper/60_cases.tex
typo

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is 0-CFA
for a small CPS language consisting of abstractions, applications, recursion and
primitive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled in different functions
(e.g., \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{compCall}, \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{compApp}, etc.).
Before refactoring into AC, these functions would take the store as a
regular argument; now every function returns a value of type \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{CompAnalysis},
which is a function (closure) value that takes and returns stores. After the
first run of analysis, we have traversed the AST and obtained a single closure
takes a store and returns a store. Therefore, the ASTs are static terms, and the
stores are dynamic terms. Now we can invoke the compiled closure with an initial
store to complete the analysis.

@@

In this section, to demonstrate the SAI approach is widely applicable, we extend
and refactor the staged abstract interpreter we presented in Section \ref{sai}
to various control-flow analysis (CFA). We first discuss calling-context
sensitive analysis for staged analyzers by tweaking allocation strategy; then we
will see how to implement store-widening with staged monads, which is a common
optimization in CFA; finally, we integrate abstract garbage collection with
staging. All of them are in fact orthogonal to staging.

@@

The abstract addresses we introduced in Section \ref{sai} is simply identical to
the variable names, which is a context-insensitive schema.
Through tweaking the address allocation strategy, we can achieve various
context-sensitive analysis \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we
demonstrate a calling-context-sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, in our staged abstract interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{KCFAAddr} is also
changed to contain a identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time with a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

@@

Every time when we call the \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{eval} function, we refresh the timestamp by calling
the \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{tick} function, which update the timestamp in the state monad. The \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{tick}
function is implemented as prepending the current expression being evaluated \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{e}
to the existing calling context, and then taking the first $k$ elements from the list.
The type \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Config} used in caching is also changed to included the timestamp.

>>>>

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is 0-CFA
for a small CPS language consisting of abstractions, applications, recursion and
primitive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled in different functions
(e.g., \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{compCall}, \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{compApp}, etc.).
Before refactoring into AC, these functions would take the store as a
regular argument; now every function returns a value of type \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{CompAnalysis},
which is a function (closure) value that takes and returns stores. After the
first run of analysis, we have traversed the AST and obtained a single closure that
takes a store and returns a store. Therefore, the ASTs are static terms, and the
stores are dynamic terms. Now we can invoke the compiled closure with an initial
store to complete the analysis.

@@

In this section, to demonstrate the SAI approach is widely applicable, we extend
and refactor the staged abstract interpreter we presented in Section \ref{sai}
to various control-flow analyses (CFA). We first discuss calling-context
sensitive analysis for staged analyzers by tweaking allocation strategy; then we
will see how to implement store-widening with staged monads, which is a common
optimization in CFA; finally, we integrate abstract garbage collection with
staging. All of them are in fact orthogonal to staging.

@@

The abstract addresses we introduced in Section \ref{sai} are simply identical to
the variable names, which is a context-insensitive schema.
Through tweaking the address allocation strategy, we can achieve various
context-sensitive analyses \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we
demonstrate a calling-context-sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, in our staged abstract interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{KCFAAddr} is also
changed to contain an identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time with a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

@@

Every time when we call the \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{eval} function, we refresh the timestamp by calling
the \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{tick} function, which updates the timestamp in the state monad. The \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{tick}
function is implemented as prepending the current expression being evaluated \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{e}
to the existing calling context, and then taking the first $k$ elements from the list.
The type \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Config} used in caching is also changed to included the timestamp.
